

# R's Popularity

R's popularity has been increasing on a daily basis.It is reported that every 1 in 100 scholarly articles indexed in Elsevier’s Scopus database 
cited R or one of its packages in 2014 [@tipmann15]. This was in 2014, by the end of 2014 there were 2925 R packages available via Comprehensive R Archive Network (CRAN), but today there are more than 10000 packages.

The number of available packages is another measure of R's popularity. Below shiny app is created to show you the number of R packages available via CRAN by now (Figure \@ref(fig:Rpacks)). At the moment you open this page, the app scans CRAN to bring you the latest number and put it in a graph. You can brush points to see further details.

```{r Rpacks, fig.cap='The number of R packages ', screenshot.opts=list(delay=100), dev='png', cache=TRUE, fig.align='center', fig.width=8, fig.height=8 , echo=FALSE}
knitr::include_app('https://burakaydin.shinyapps.io/Rpacks/', height = '550px')
```


Even though the increase in the number of packages is a measure of R's popularity, it is more important to know if people are using these packages. Another shiny app is created by David Robinson, with his kind permission, Figure \@ref(fig:Rpackdown) can tell you the number of downloads for each package when the package name is provided.

```{r Rpackdown, fig.cap='The number of R packages downloads ', screenshot.opts=list(delay=10), dev='png', cache=TRUE, fig.align='center', fig.width=8, fig.height=8, echo=FALSE}
knitr::include_url('https://dgrtwo.shinyapps.io/cranview/', height = '900px')
```



__There are other indicators for R's popularity.__

1. R has become a universal language for data analysis and it offers new methods sooner  [@Muenchen11].

2. Among the Institute of Electrical and Electronics Engineers (IEEE) community, it is reported to be one of the top 5 programming languages <a href="http://blog.revolutionanalytics.com/popularity/" target="_blank">see</a>

3. R courses are offered in universities and as Massive Open Online Courses.

4. Private companies use R, i.e. Google, Twitter, Microsoft.