

# Basics {#basics}

R can create advanced outputs. This section aims to familiarize readers with the basic principles of R before creating or reproducing advanced outputs.

## Functions

A programmable calculator enables its users to write and save functions. It is useful for an R user to understand how R functions work.

### R as a Basic Calculator.
R can calculate. See the examples below followed by R syntax.

\begin{equation}
1+1=2 
(\#eq:bc1)
\end{equation}

```{r, collapse=T}
1+1
```

\begin{equation}
1-1=0 
(\#eq:bc2)
\end{equation}

```{r,collapse=T}
1-1
```

\begin{equation}
1 + (2 / 3) - (2 * 6.5) = -11.33 
(\#eq:bc3)
\end{equation}

```{r, collapse=TRUE}
1 + (2 / 3) - (2 * 6.5)
```

\begin{equation}
sin(30) + 4^3 + log(4) + e^3 + \sqrt{7} = 87.13 
(\#eq:bc4)
\end{equation}

```{r, collapse=TRUE}
sin(30) + 4^3 + log(4) + exp(3) + sqrt(7)
```

When typed and run with R, Equations \@ref(eq:bc1) through \@ref(eq:bc4) are calculated but not kept. If an outcome of an R operation will be used later, it should be named. When a name is assigned, the outcome can be recalled easily. The assigned outputs are saved in the R environment throughout the session.  Assignment can be done with "=", "<-" or "<<-". This book uses "=".
Let's assign a name for the Equations \@ref(eq:bc1) through \@ref(eq:bc4)'s outputs.

```{r, collapse=TRUE}
a=1 - 1
b=1 + 1
c=1 + (2 / 3) - (2 * 6.5)
d=sin(30) + 4^3 + log(4) + exp(3) + sqrt(7)
```

It is possible to operate with these assigned variables. 
```{r, collapse=TRUE}
a+b+c+d
```

It is possible to overwrite.
```{r, collapse=TRUE}
e=3+2
e
e=e+10
e
```

It is possible to rename. (Note: R is case sensitive).
```{r, collapse=TRUE}
Equation1_output=a
Equation1_output + b + c + d  #is equal to a+b+c+d
```


### R as a Programmable Calculator

A function basically has 3 parts, an input, a process and an output. Let's use an analogy, assume that below functions are created by a teacher to examine test scores.

#### Single input - Single output  

A simple function is given below and named as _constant5_. Let's assume it adds 5 points to each score. The _constant5_ function takes a value, adds 5 and produces an output.

```{r,collapse=T}
constant5=function(input){
    output=input+5
    return(output)
    }

constant5(input=50)
constant5(100)
constant5(120)
```

With above code, we use R as a programmable calculator. We define  _constant5_ as a _function_ that takes an input, processes it by adding 5 _(input+5)_, creates an output _(output=input+5)_ and reports it _(return(output))_.  All these steps should be given in _{   }_. 


Another simple function will be that _systematic1_, adds 1% for each score. It will take a value, add 1% and produce an output.

```{r,collapse=T}
systematic1=function(input){
    output=input+(input/100)
    return(output)
    }

systematic1(input=50)
systematic1(100)
systematic1(120)
```


#### Multiple input - Single output 

Above two examples use one single value as an input. Let us use two values for _nomistake_ function. In this example, let`s say the teacher cuts 0.2 points for each spelling mistake. For example, if a grade is 90, it will go down to 88.8 if there are 6 spelling errors. The _nomistake_ asks for a grade and the number of spelling errors to calculate the reduced grade.

```{r, collapse=TRUE}
nomistake=function(grade, nserror){
    output=grade - (0.2 * nserror)
    return(output)
    }
nomistake(grade=90,nserror=6)
nomistake(90,17)
```

Inputs for an R function are generally called _arguments_. _nomistake_ is programmed to receive 2 arguments to calculate one single output. It is possible to create functions with multiple arguments and multiple outcomes.

#### Multiple input - Multiple output 

The _feedback_ function asks for number of correct responses and points for each to calculate a total score. It also provides the number of correct responses needed for a full score of 100.

```{r, collapse=TRUE}
feedback=function(correct, point){
    total=correct*point
    remained=(100-total)/point
    output=c(paste("score:", total," missed items:",remained))
    return(output)
    }
feedback(correct=20,point=2)
feedback(27,2)
```

#### Basic error

R functions need arguments to work. Please see the following error if you forget to feed _point_ parameter into  the _feedback_ function 

```{r, collapse=TRUE,error=T}
feedback=function(correct, point){
    total=correct*point
    remained=(100-total)/point
    output=c(paste("score:", total," missed items:",remained))
    return(output)
    }
feedback(correct=20)

```

#### Basic warning
R functions can produce warnings. Let us create _nomistake2_ that calculates the remaining score after cuts

```{r, collapse=TRUE,error=T}
nomistake2=function(grade, nserror){
    output=grade - (0.2 * nserror)
    return(output)
    }
nomistake2(grade=50,nserror=10)
```

We can produce a warning if the final score is lower than 0.

```{r, collapse=TRUE}
nomistake2=function(grade, nserror){
      output=grade - (0.2 * nserror)
      if (output<0) 
    warning("Final score is lower than 0")  
    return(output)
    }
nomistake2(grade=10,nserror=60)
```


#### Basic failure
A function can stop. Let us create _nomistake3_ that calculates the final score. However,this time, it stops if the score is lower than 20 to avoid further cuts. 
```{r, collapse=TRUE,error=T}
nomistake3=function(grade, nserror){
 
    if ((grade)<(20)) 
    stop("Score is already low")
  
     output=grade - (0.2 * nserror)
      return(output)
    }
nomistake3(10,9)

```


### Help!
Although applied R users do not need to write new functions, they should know the principles of how R functions work. Whenever an R function throws a warning or an error, it generally is caused by the users (or their data) rather than the function itself. 

R basically runs on functions. Researchers write functions, place them in R packages and make them available. There are currently 10000+ R packages available via Comprehensive R Archive Network. R version 3.3.1 downloads to your computer with 30 packages that includes thousands of functions.  

One of the main packages that have been downloaded to your computer is called _base_, and it has 1200+ functions. For example this package has the _mean_ function to calculate the arithmetic mean. Packages and functions are generally well documented. Users should effectively use the documentations via _help_ function, _?_ or _??_. _example_ functions may also be helpful.

```{r, collapse=T, eval=F}
help("base")  # see description, you can click on index at the bottom to see 1200+ functions
help(mean)    # see the mean function and its arguments
?mean         # see the mean function and its arguments
??mean        # see the mean function and its arguments
example(mean) # see an example

```


## R Data Types
Vectors, matrices, variable types, factors, missing values and data frames are briefly introduced.

### Vectors
R can create vectors using _c()_ function. Let's create grades for 10 students
```{r, collapse=T}
grades=c(40,50,53,65,72,77,79,81,86,90)
grades
```

R can operate with vectors.

```{r, collapse=T}
grades=c(40,50,53,65,72,77,79,81,86,90)
grades+10
grades+(grades*0.10)
grades*grades
grades2=c(30,40,46,58,64,66,69,72,74,81)
(grades+grades2)/2
grades*0.4 + grades2*0.6


```

There are useful functions to create vectors. For example the _rep_ function (see example(rep)) is helpful to repeat values.  

The _rnorm_ function can create random variables. If you run _?rnorm_ you will see it has three arguments, _rnorm(n, mean = 0, sd = 1)_ . This function requires the number of observations ( _n_ ) argument to be provided. By default the mean is set to be 0 and standard deviation to be 1. However, you can change the default for example by running _rnorm(12,mean=10,sd=2)_ to create 12 observations from a normal distribution with mean 10 and standard deviation 2. A similar function is _runif(n, min = 0, max = 1)_ to generate _n_ observations from a uniform distribution on the interval from minimum=0 to maximum=2. You can change the interval, for example by running _runif(12, min = 10, max = 37)_.

```{r, collapse=T}
a=1:12            # a is a regular sequence from 1 to 12 created with ':'
rep(0,12)         # repeat zero 12 times
rep(1:5,each=3)   # repeat 1 to 5 each 3 times 
rep(1:5,times=3)  # repeat 1 to 5 , 3 times
seq(from=1,to=12) # create 1 to 12 sequence
seq(1,25,by=2)    # create 1 to 25 by 2
seq(1,6,by=0.5)   # create 1 to 6 by 0.5
rnorm(12)        # create  12 random observations from ~N(0,1)
rnorm(12,mean=10,sd=2) #create 12 random observations from ~ N(10,2)
runif(12, min = 10, max = 37) # create 12 random observations from a uniform distribution.
```



### Matricies
R can create matrices and operate.

```{r, collapse=T}
A=matrix(1:16,ncol=4,nrow=4) #create a 4 x 4 matrix
A
B=matrix(runif(16,min=20,max=40),ncol=4) #create a 4 x 4 matrix

# example operations
A+B    # add
A*B    # multiply
A%*%B  # matrix multipication
t(B)   # transpose
```

### Variables {#variables}
It is important to know the data before running basic or sophisticated analyses. In an R environment, a variable subject to an analysis is generally defined as nominal, ordered, continuous, missing or date variable.

#### Nominal
In R, a nominal variable can be represented alphanumerically. However the interpretation of a nominal variable is not numeric. It is helpful for naming a characteristic rather than quantifing it. Below commands can create nominal vectors.

```{r, collapse=T,cache=T}
address=c("AAX","BBZ","CBT","DBA","DDC","XZT")
gender=c("M","F","F","M","F","M")
id=sample(letters,6)
treatment=rep(c("cntrl","trt"),each=3)
city=as.character(1:6)
```

#### Ordered
An ordered variable includes more information compared to a nominal variable. It represents order but the difference between values is not informative. Below commands can create ordered variables. The _level_ argument for an ordered factor provides the information of order. If the _level_ argument is not provided, R , by default, sorts the unique set of given values into increasing order.

```{r, collapse=T,cache=T}
item1=ordered(c("poor","average","good","good","poor","poor"),
              levels=c("poor","average","good"))
ses=ordered(c(1,3,2,2,1,3),levels=c("1","2","3"))
```

#### Continuous
An interval or ratio (true-zero variable) provides more information compared to ordinal and nominal variable. The difference between values is informative. Below commands can create continuous variables.
```{r, collapse=T,cache=T}
grade=c(52,75,39,62,24,86)
score=rnorm(n=6,mean=160,sd=5)
```


#### Date Variable
One of the several date variable creation methods is using the as.Date() function. It will try to convert what is provided into a date. This is a flexible function and you can use the _format_ argument to provide the information on how you enter a date. By default it looks for a format of _YYYY-MM-DD_ .Another convenient way to input a date variable might be in _MM/DD/YYYY_ format. This is possible by using _format="%m/%d/%y"_ . _Sys.Date()_ function will give you today's date in a _YYYY-MM-DD_ format. You can operate with dates, for example the _Sys.Date( )-birthday_ command below calculates the number of days between the provided birthdays and today.

```{r, collapse=T,cache=T}
birthday=as.Date(c("1984-06-01","1988-10-20","1990-12-01",
                   "1978-03-23","1974-08-22","1994-11-04"))

birthday

holidays=as.Date(c("01/01/2016","04/23/2016","05/19/2016","08/30/2016","09/29/2016"),
                 format="%m/%d/%y")

holidays

Sys.Date( )
Sys.Date( )-birthday
```

#### Logical variable
A logical variable takes a value of either TRUE or FALSE. When forced to be numeric, a logical variable takes the form of 1 and 0. Below command tests the grade variable whether its elements are larger than its mean or not. 

```{r, collapse=T,cache=T}
grade=c(52,75,39,62,24,86)    # create grades
grade>mean(grade)             # create TRUE-FALSE by testing if the grade is larger than the mean
as.numeric(grade>mean(grade)) # force the logical variable to be 1 and 0.
```

### Factors
R has a data type of _factor_. It can be considered as a general frame for nominal and ordered variables. 
```{r, collapse=T,cache=T}
course=factor(c("Cook","Plumber","Designer","Plumber","Cook","Plumber"))
ga1=factor(c(1,1,3,4,2,3),levels = 1:4,
           labels=c("StronglyDisagree","Disagree","Agree","StronglyAgree"))
ga2=factor(c(1,3,4,4,2,3),ordered = T)
ga3=gl(n=3,k=2,labels=c("A","B","C"),ordered=F)

```

Factors are important data types. Levels of a factor should be examined. It might be necessary to drop levels if they are not used in the variable. 
For example, if the main data has a factor , lets say _Color_ and the levels are "blue", "green" and "yellow".  Assume a subset is chosen from the data and it has only "blue" and "yellow", R will still treat it as factor with 3 levels.This will cause problems. 

The droplevel function drops unused levels. Examine the code below;
```{r, collapse=T,cache=T}
#the ga4 factor is defined with 4 levels A,B,C and D.
#BUT the data has only 1s,2s and 3s, the level D is not used.
ga4=factor(c(1,1,3,2,2,3),levels = 1:4,labels=c("A","B","C","D"))
ga4
droplevels(ga4)
```


### Missing Values
The data might be incomplete. R uses __NA__ (not available) to represent missing values. 
```{r, collapse=T,cache=T}
incomeSource=c("wage","wage","pension",NA,NA,"wage")
houseMember=c(3,2,3,NA,NA,4)
```

NOTE: Missing data indicators might be confusing. Notice the difference between NA, <NA> , " " (empty cell) and a predefined missing indicator, such as -99.

```{r, collapse=T,cache=T}
temp = factor(c('wage','pension', NA, 'NA'," ",-99,"-99"))

#for a factor or a character variable NA is shown as <NA> to specify a true missing cell
# NA without < > represents a factor level
# " " also represents a factor level
#-99 and "-99" represents the same factor level

is.na(temp) # identifies only the third element as a missing value.

#A possible solution
temp[temp=='NA' | temp==" "| temp== -99 | temp== "-99"]=NA

#check
is.na(temp)

#DO NOT forget to drop levels
temp=droplevels(temp)

```

### Data Frames {#dataframes}
A data frame includes variables. Assuming a social scientist is generally interested in the relationships between the variables, a data frame is their main R structure. Below command creates a data frame using some of the earlier created variables. 

```{r, collapse=T,cache=T}
# reminder
# id=sample(letters,6)

# treatment=rep(c("cntrl","trt"),each=3)

# gender=c("M","F","F","M","F","M")

# item1=ordered(c("poor","average","good","good","poor","poor"),
#              levels=c("poor","average","good"))

# ses=ordered(c(1,3,2,2,1,3),levels=c("1","2","3"))

# grade=c(52,75,39,62,24,86)

# incomeSource=c("wage","wage","pension",NA,NA,"wage")

# birthday=as.Date(c("1984-06-01","1988-10-20","1990-12-01",
#                   "1978-03-23","1974-08-22","1994-11-04"))

# course=factor(c("Cook","Plumber","Designer","Plumber","Cook","Plumber"))

basic_data=data.frame(id,treatment,gender,item1,ses,
                      grade,incomeSource,birthday,course)
basic_data

```

Data can be entered manually into R. However this is generally not the case. When data are transferred into the R environment, a useful function to check its internal structure is named as _str_.

```{r, collapse=T,cache=T}
str(basic_data)
```

## R Packages
R version 3.3.1 downloads to a computer with 30 packages that includes thousands of functions.These packages are stored under _system library_. Other useful functions are created by R users and made available to R community. For example linear mixed effect models can be analyzed R using _lme4_[@lme4pack] package. This package is cited more than 1500 times and has been downloaded more than 60000 times. You can use Figure \@ref(fig:Rpackdown) to check its current usage. R packages are generally available via CRAN and they are generally not archived further if they are maintained properly. You can download R packages into your computer and store them locally, under _user library_ . You need to load (activate) the packages in each session before you can use them. 

You have probably noticed that R, RStudio and R packages are interconnected. When you download Rstuido after you have downloaded R, the Rstudio scans your computer, locates R and connects to it. Both R and RStudio can locate your libraries unless you manually manipulated the file locations. If you wonder the location of your R packages you can run _.libPaths()_ function. 

R packages located in CRAN can easily be downloaded into your machine using RStuido's __Packages__ tab, or you can directly type _install.packages("packagename")_. When you open a new session, some of the main packages are loaded automatically. When a package is loaded, you can see the tick in the _Packages_ tab. If the package you plan to use in a session is not loaded, you can click the box or you can directly type _library("packagename")_. You can see these steps in (Video4 \@ref(fig:Video4)).

```{r Video4, fig.cap='Download R Packages', fig.width=8, fig.height=6 , echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/BiJhU0Rn6t4")
``` 


## The Workspace {#theworkspace}
When a session is started by opening an R script, every operation takes place in the working space. Every step is recorded and can be seen in _History_ tab of R Studio. Working space can be saved when closing the session. The objects created in a session are kept in the space. You can use _ls()_ function to see your objects in the working space, you can also check _Environment_ tab of R Studio.

The objects in a workspace can easily be saved into the working directory as separate outputs. Also the objects in the working directory can easily be loaded into the working space. Here _easily_ refers to the unnecessity of providing a path. If the path is provided, you can save or load objects from different directories.

You can run _getwd()_ command to see your working directory. You can change the working directory within a session using _setwd()_ function. Alternatively you can change the directory using _Session_ tab of R Studio. The data input and output is covered more broadly in the next chapter.
