

#  Data Sets

In section \@ref(dataframes) we have illustrated how to enter data manually. However the data are generally available beforehand. This section aims to illustrate how to (a) import data into R, (b) perform basic data manipulation  and (c) export data.


##Import Data
A data set might be available in different formats. Some of the most commonly used formats are .csv, .sav, .Rdata, .txt. Importing data and checking their status are initial steps and should be conducted carefully. As mentioned in \@ref(theworkspace), if the data and the R script are in the same folder, a detailed path to locate the data is NOT needed. 

### CSV
CSV stands for comma separated values. Microsoft Excel is a useful tool to create this format. The csv is simpler compared to xls, xlsx, xlsb or other Excel formats. A csv file can be imported into R working space using _read.csv_ function. Following code is the simplest specification to read a csv file

```{r, collapse=T, eval=F}
data=read.csv("dataname.csv")   # works if dataname.csv is located within the working directory

#In a windows machine
data=read.csv("C:\\Users\Desktop\\folderX\\data.name.csv")  # with path
data=read.csv("C:/Users/Desktop/folderX/data.name.csv")  # with path
#NOTE: A backslash (\) will cause error in a windows machine 

```

Use _?read.csv_ to view its arguments. See following notes for relatively important arguments

a) _header=FALSE_ or _header=TRUE_ indicates whether the file contains variable names.    
b) _na.strings_ is used for declaring missing value indicators. This is generally critical. For example _na.strings = "-99"_ indicates that -99s should be interpreted as _NA_ or _na.strings = c("-99" , "-9" )_ indicates that both -99s and -9s represents missing data.
c) _stringsAsFactors=TRUE_ or _stringsAsFactors=FALSE_ indicates whether character vectors should be converted into factors.
d) _col.names_ allows renaming variables while reading the data. For example, when reading three variables, _col.names_ should have three elements, such as _col.names=c("A1","B2","C3")_.



Use _read.csv2_ function when decimals indicated with a comma and values are separated with semicolon. Alternatively use _sep=";"_ and _dec=","_ arguments in _read.csv_ function. 

A silent video is recorded (Video5 \@ref(fig:Video5)) to show these steps.

```{r Video5, fig.cap='Read CSV file', fig.width=8, fig.height=6 , echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/RjiKIWxaZBk")
```


### SPSS
SAV files are common data formats, at least among the social scientists. The package _foreign_ [@foreign] includes _read.spss_ function.
```{r, collapse=T, eval=F}
require(foreign)
?read.spss
data=read.spss("dataname.sav",to.data.frame=TRUE)   
# works if dataname.sav is located within the working directory

```

### Rdata
Rdata format is a memory friendly alternative. By default, it includes a name for the data set. 

```{r, collapse=T, eval=F}
load("dataname.Rdata")  # works if dataname.Rdata is located within the working directory
```

### Pull online {#pullonline}

Its possible to import data from the web, but these procedures are well beyond the scope of this material. The basic approach includes mainly three steps, (a) correctly specify the location , (b) correctly specify the data format, (c) download and import or directly import into R. Following code should import the World Bank data introduced in Section \@ref(dataWBT).

```{r, collapse=T, eval=F}
#read csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
dataname=read.csv(urlfile)
str(dataname)


# load Rdata from an online repository
urlfile2='https://github.com/burakaydin/materyaller/blob/gh-pages/ARPASS/dataWBT.Rdata?raw=true'
load(url(urlfile2))
str(dataWBT)

```

These data sets can be downloaded as a file from the <a href="https://github.com/burakaydin/materyaller/tree/gh-pages/ARPASS" target="_blank">Github Repository</a>. Alternatively, an excel file is located <a href="https://www.aydinburak.net/sarp" target="_blank">here</a>.

### Read data through R studio
When the data are not located within the working directory or point-click approach is preferable, _Import Dataset_ gadget located in _Environment_ window might be helpful. A silent video is recorded (Video6 \@ref(fig:Video6)) to show how to read a csv file through Rstudio. It is also possible to read from Excel, SPSS, SAS and Stata.

```{r Video6, fig.cap='Read CSV file through R Studio', fig.width=8, fig.height=6 , echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/D_5wJFmlmxk")
```


## Basic Data Manipulation

Generally a data set should be processed after it is imported. This section illustrates basic manipulation procedures, (a) replace values, (b) subsetting, (c) create new variables, (d) reshape (e) convert between variable types, (f) delete cases.

### Replacing values
It is possible to replace a specific datum or a value. It is also possible to rename variables. The R package _plyr_ [@plyrpack] might be useful.

```{r, collapse=T, eval=F, cache=T,message=F}

# load csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
tempdata=read.csv(urlfile)

#remove URL 
rm(urlfile)


# replace a specific datum:i.e. row 151 column 16 should be 30 
tempdata[151,16]=30

# replace a specific datum using locaters
# row 151 belongs to id 67034022 and column 16 is named "age".
tempdata[tempdata$id==67034022,"age"]=32

# Replace values
# Replace treatment values
# originally has 1 for treatment and 2 for control
# Replace 1s with "trt" and 2s with "cnt" 
tempdata[tempdata$treatment==1,"treatment"]="trt"
tempdata[tempdata$treatment==2,"treatment"]="cnt"

# Or use ifelse function
# replace "wage01" If "wage01" equals "Yes" change to 0.5, otherwise -0.5
tempdata$wage01=ifelse(tempdata$wage01=="Yes",0.5,-0.5)

# Or use mapvalue function in the plyr package
require(plyr)
# create a new variable named pension01NEW, 0 if pension01 is "No", 1 if "Yes"
tempdata$pension01NEW <- mapvalues(tempdata$pension01, 
                                   from=c("Yes","No"),to=c("1","0"))


#rename a variable
#rename 4th and 5th column
colnames(tempdata)[4]="course"
colnames(tempdata)[5]="region"

#rename at the same time
colnames(tempdata)[c(17,21)]=c("Tinc","WAGE1")

#rename using plyr package
tempdata <- rename(tempdata,c('gen_att'='GENDERATT'))

#use head(tempdata) to visually check
#use summary(tempdata) to check

#remove tempdata
rm(tempdata)

```



### Subsetting

It is easily possible to subset a data set.


```{r, collapse=T, eval=F, cache=T,message=F}

# load csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
tempdata=read.csv(urlfile)

#remove URL 
rm(urlfile)


# select only city=ISTANBUL
istDAT=tempdata[tempdata$city=="ISTANBUL",]

# select only city=ISTANBUL and first 8 columns
istDAT18=tempdata[tempdata$city=="ISTANBUL",1:8]

# select only city=ISTANBUL and Gender Attitude score larger than 2
istDATGAT2=tempdata[tempdata$city=="ISTANBUL" | tempdata$gen_att >2 ,]

# Alternatively use the subset function
# use the select argument to specify the columns, otherwise all columns are selected
istDATGAT2B=subset(tempdata, city=="ISTANBUL" | tempdata$gen_att >2, select=1:8) 

#subset based on variable values
item1_123 <- tempdata[tempdata$item1 %in% c(1,2,3), ]

#remove all objects in workspace
rm(list=ls())

```


### Creating new variables
Procedures to create new variables are covered in Section \@ref(basics). Following computations are mainly a review. 

```{r, collapse=T, eval=F, cache=T,message=F}
# load csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
tempdata=read.csv(urlfile)

#remove URL 
rm(urlfile)

# create sum for item2 to item6
tempdata$itemSUM=with(tempdata,item2+item3+item4+item5+item6)

# create average of item2 to item6 (notice na.rm)
tempdata$itemAVE=with(tempdata,
                      rowMeans(cbind(item2,item3,item4,item5,item6),na.rm=T))

#or
tempdata$itemAVE=rowMeans(tempdata[,10:14],na.rm = T)

# Create average by city
tempdata$CityAVEscore =with(tempdata, ave(itemAVE,city,FUN=function(x) mean(x, na.rm=T)))

#or
tempdata=merge(tempdata, aggregate(itemAVE ~ city, data = tempdata, FUN=mean, na.rm=TRUE), 
          by = "city", suffixes = c("", "citymean"),all=T)

#or take each item's average by city
tempdata=merge(tempdata, aggregate(cbind(item2,item3,item4,item5,item6) ~ city, 
                                   data = tempdata, FUN=mean, na.rm=TRUE), 
                                   by = "city", suffixes = c("", "Citymean"),all=T)


# categorize variables. Create 0s if itemAVE is lower than 2 and 1 otherwise
tempdata$itemAVE01=ifelse(tempdata$itemAVE<2,0,1)

# create 1s if the average is between 0 and 1.8
# create 2s if the average is between 1.8 and 2.5
# create 3s if the average is between 2.5 and 5
tempdata$itemAVE123=with(tempdata,cut(itemAVE, breaks=c(0,1.8,2.5,5), labels = FALSE))
# check right=TRUE argument.
# i.e if right=T  values exactly equal to 1.8 goes into category 1
#     if right=F  values exactly equal to 1.8 goes into category 2
     

```



### Reshaping data
It might be needed to reshape the data. Mainly from a wide format to long format or vice versa. The R package _tidyr_ [@tidyrpack] might be useful.

```{r, collapse=T, eval=F, cache=T,message=F}
# load csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
tempdata=read.csv(urlfile)

#remove URL 
rm(urlfile)

# from wide to long. Create one single item column based on item1 to item6
# while keeping other values in the data set 
library(tidyr)
data_long = gather(tempdata, item, score, item1:item6, factor_key=TRUE)

#sort data by id
data_long=data_long[order(data_long$id),] 

# from long  to wide.
data_wide = spread(data_long, item, score)

## remove all objects but certain in workspace
rm(list=setdiff(ls(),c("tempdata")))

```

### Converting between variable types
Converting numeric values to factors or vice versa might be needed.


```{r, collapse=T, eval=F, cache=T,message=F}
# load csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
tempdata=read.csv(urlfile,stringsAsFactors = F)

#remove URL 
rm(urlfile)

#check treatment variable's structure 
str(tempdata$treatment)

#convert numeric to factor.
tempdata$treatmentFactor=factor(tempdata$treatment,labels=c("treatment","control"))

#convert factors to numeric (when numbers show up as a character variable)

#create illustrative variable
tempdata$iv1=factor(rep(c("1","2","3"),length=nrow(tempdata)))
tempdata$iv1numeric=as.numeric(levels(tempdata$iv1))[tempdata$iv1]
#or
tempdata$iv1numeric=as.numeric(as.character(tempdata$iv1))

#convert NAs to -99  
tempdata[is.na(tempdata)]= (-99)

#remove all objects in workspace
rm(list=ls())
```


### Delete cases

It might be needed to delete an element, an entire row or an entire column

```{r, collapse=T, eval=F, cache=T,message=F}
# load csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
tempdata=read.csv(urlfile,stringsAsFactors = F)

#remove URL 
rm(urlfile)

#Delete a specific element i.e row 3 column 5
tempdata[3,5]=NA

#Delete a row. i.e the 3rd 
tempdata[3,]= NA

#or
tempdata=tempdata[-3,]

#delete a column, i.e the 5th 
tempdata$course_taken=NULL

#listwise deletion, remove any row with missing data
# subset temp for illustrative purposes
temp=tempdata[,1:10]

# listwise deletion
temp=na.omit(temp)

#remove all objects in workspace
rm(list=ls())

```



## Export Data

It is possible to export R objects. Unless the working directory is changed using _setwd()_ or a path is provided, all the objects are sent to your current working directory.

```{r, collapse=T, eval=F, cache=T,message=F}
# load csv from an online repository
urlfile='https://raw.githubusercontent.com/burakaydin/materyaller/gh-pages/ARPASS/dataWBT.csv'
tempdata=read.csv(urlfile,stringsAsFactors = F)

#remove URL 
rm(urlfile)

#create objects
# select rows 1 to 20 and columns 1 to 5
subset1=tempdata[1:20,1:5]
object2=mean(tempdata$item1,na.rm = T)


#Check working directory
getwd()

# save as an R object
save(subset1,file="subset1Rfile.Rdata")
# provide a path if needed
save(object2,file="C:/Users/Desktop/object2Rfile.Rdata")

# Export as a csv
write.csv(subset1,file="subset1CSVfile.csv",row.names = F)

#Export to SPSS
library(foreign)
write.foreign(subset1, "subset1SPSfile.txt","subset1SPSfile.sps", package="SPSS")

#remove all objects in workspace
rm(list=ls())

```
